---
title: "R Lab. - Exercise 1"
author: "Michele Guadagnini - Mt. 1230663"
date: "April 15, 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1 - Vectors and dataframes

Importing scottish lakes data:

```{r scottish lakes data, results='asis'}
Names <- c("Ness", "Lomond", "Morar", "Tay", "Awe", "Maree", "Ericht", "Lochy", "Rannoch", 
           "Shiel", "Katrine", "Arkaig", "Shin")
Vols  <- c(7.45, 2.6, 2.3,  1.6,  1.2, 1.09, 1.08, 1.07, 0.97, 0.79, 0.77, 0.75, 0.35)#km3
Area  <- c(56,   71,  27,   26.4, 39,  28.6, 18.6, 16,   19,   19.5, 12.4, 16,   22.5)#km2
Lengs <- c(39,   36,  18.8, 23,   41,  20,   23,   16,   15.7, 28,   12.9, 19.3, 27.8)#km
Maxdp <- c(230,  190, 310,  150,  94,  114,  156,  162,  134,  128,  151,  109,  49  )#m
Meandp<- c(132,  37,  87,   60.6, 32,  38,   57.6, 70,   51,   40,   43.4, 46.5, 15.5)#m

scottish.lakes <- data.frame(Names, Vols, Area, Lengs, Maxdp, Meandp)
colnames(scottish.lakes) <- c("Name","Volume [km³]","Area [km²]","Length [km]",
                              "Max dp [m]","Mean dp [m]")
knitr::kable(scottish.lakes, caption="Scottish Lakes data frame")
```

### 1) Evaluate the highest and lowest volume and area lake

```{r volume and area extremes}
maxvol <- max(scottish.lakes$Volume, na.rm=TRUE)
minvol <- min(scottish.lakes$Volume, na.rm=TRUE)
maxn <- scottish.lakes$Name[which.max(scottish.lakes$Volume)]
minn <- scottish.lakes$Name[which.min(scottish.lakes$Volume)]

message(paste("Maximum volume is", maxvol, "km³ of Loch", maxn, 
              "\nMinimum volume is", minvol, "km³ of Loch", minn))

maxarea <- max(scottish.lakes$Area, na.rm=TRUE)
minarea <- min(scottish.lakes$Area, na.rm=TRUE)
maxn <- scottish.lakes$Name[which.max(scottish.lakes$Area)]
minn <- scottish.lakes$Name[which.min(scottish.lakes$Area)]

message(paste("Maximum area is", maxarea, "km² of Loch", maxn,
              "\nMinimum area is", minarea, "km² of Loch", minn))
```

### 2) Order the frame with respect to the area and determine the two largest area lakes

```{r largest area lakes}
scottish.lakes <- scottish.lakes[order(-scottish.lakes$Area),]
lan <- scottish.lakes$Name[1:2]
la  <- c(scottish.lakes$Area[1:2])

message(paste("The lake with largest area is Loch", lan[1], "with", la[1], "km²" , 
              "\nThe lake with second largest area is Loch", lan[2], "with", la[2], "km²"))
```

### 3) By summing up the areas occupied by the lakes, determine the area of Scotland covered by water

```{r total area}
WaterArea <- sum(scottish.lakes$Area)
message(paste("Total surface covered by water is", WaterArea, "km²"))
```

\newpage
## Exercise 2 - DAAG and Tibble

Importing needed packages:
```{r packages, message=FALSE, warning=FALSE}
#install.packages(c('DAAG','tibble'), type='source')
library(DAAG, tibble)
#library(help=DAAG)
library(tidyverse)
```

Loading the Australian athletes data frame:
```{r ais, results='asis'}
data(ais)
#?ais
tbais <- tibble(ais)
knitr::kable(tbais[1:5,], caption="Australian athletes data frame")
```

### 1) Create a table grouping the data by gender and by sport; produce a barplot with the table adding a legend

```{r table}
tbl <- table(tbais$sex, tbais$sport)
tbl
```

```{r barplot, echo=TRUE}
maxY <- max(tbl)
barplot(tbl, beside=TRUE, col=c("mistyrose", "lightblue"), 
        legend=rownames(tbl), las=2, ylim=c(0,maxY+5))
title( main="Australian athletes" )
```

### 2) Determine if any of the columns holds missing values

```{r nacheck}
any(is.na(tbais))
```

### 3) Produce boxplots of the main blood variables (’red blood cell counts’, ’white blood cell counts’, ’hematocrit’ and ’hemaglobin concentration’), for different kind of sports

```{r boxplots, fig.height=6}
# custom colors list
cols <- c("red","orchid1","purple","blue","cyan","green","gold1",
          "darkorange1","sienna","forestgreen")

par(mfrow=c(2,2))
boxplot(rcc~sport, data=tbais, main='Red Cell Counts (RCC) by sport', xlab='Sport',
        ylab='RCC', col=cols, medcol='white', las=2)
boxplot(wcc~sport, data=tbais, main='White Cell Counts (WCC) by sport', xlab='Sport',
        ylab='WCC', col=cols, medcol='white', las=2)
boxplot(hc~sport, data=tbais, main='Hematocrit by sport', xlab='Sport', 
        ylab='Hematocrit', col=cols, medcol='white', las=2)
boxplot(hg~sport, data=tbais, main='Hemaglobin concentration by sport', xlab='Sport',
        ylab='Hemaglobin', col=cols, medcol='white', las=2)
```

### 4) Make some scatter plot correlations of the same blood variables using different colors and symbols for the two genders in the sample

```{r scatterplots, fig.height=6}
vars <- list(tbais$rcc, tbais$wcc, tbais$hc, tbais$hg)
keys <- list("RCC", "WCC", "Hematocrit", "Hemaglobin")

pairs(vars, keys, main="Correlations scatter plots matrix", 
      col=ifelse(tbais$sex=='f', "coral2", "blue"), 
      pch=ifelse(tbais$sex=='f', 24, 25),
      bg =ifelse(tbais$sex=='f', "mistyrose", "lightblue"),
      lower.panel=NULL,
     )
par(xpd=TRUE)
legend(0.1,0.3, legend=unique(tbais$sex), 
       col = c("coral2", "blue"),
       pt.bg = c("mistyrose", "lightblue"),
       pch = c(24, 25),
       title="Legend:"
      )
```

\newpage
## Exercise 3 - COVID-19

```{r packages-data, warning=F, message=F}
needed_packages <- c('lubridate', 'readxl', 'curl')
already_installed <- needed_packages %in% installed.packages()
for (pack in needed_packages [!already_installed]) {
    message(paste("To be installed : ", pack, sep = " "))
    install.packages(pack)
}
library(lubridate)
library(readxl)
library(curl)
url <- "https://www.ecdc.europa.eu/sites/default/files/documents/"
fname <- "COVID-19-geographic-disbtribution-worldwide-"
date <- lubridate::today() - 1
ext = ".xlsx"
target <- paste(url, fname, date, ext, sep = "")
message("target :", target)
tmp_file <- tempfile("data", "/tmp", fileext = ext)
tmp <- curl::curl_download(target, destfile = tmp_file)
covid <- readxl::read_xlsx(tmp_file)
```

### 1) Exploring the structure of the loaded data frame:
```{r covidtibble}
summary(covid)
#excluding last column (popData2018) due to page size limit
knitr::kable(covid[order(-covid$year, -covid$month, -covid$day),][1:10,1:9],
             caption="Head of covid data frame")
```

### 2) Selecting yesterday data with more cases or more deaths

```{r yest}
yest <- Sys.Date()-1
covidYest <- covid[covid$dateRep==yest,]

covidYestNewCases <- covidYest[covidYest$cases > 200,]
x <- matrix( c(covidYestNewCases$cases, covidYestNewCases$deaths), ncol=2)
colnames(x) <- c("cases", "deaths")
rownames(x) <- covidYestNewCases$countriesAndTerritories
tblCases <- as.table(x)
tblCases

covidYestNewDeaths <- covidYest[covidYest$deaths > 200,]
y <- matrix( c(covidYestNewDeaths$deaths, covidYestNewDeaths$cases), ncol=2)
colnames(y) <- c("deaths", "cases")
rownames(y) <- covidYestNewDeaths$countriesAndTerritories
tblDeaths <- as.table(y)
tblDeaths
```

### 3) Selecting top ten countries in term of cases

```{r topten}
totCases  <- aggregate(covid$cases, by=list(covid$geoId), FUN=sum)
totCases  <- totCases[order(-totCases$x),][1:10,]
TopgeoIds <- list(totCases[[1]])

covidTop10 <- covid[covid$geoId %in% TopgeoIds[[1]],]
Toplist    <- group_split(covidTop10 %>% group_by(geoId))

for (i in c(1:10)) {    
    Toplist[[i]] <- Toplist[[i]][order(Toplist[[i]]$year, Toplist[[i]]$month,
                                       Toplist[[i]]$day),]
    Toplist[[i]][, "cumcases"]  <- cumsum(Toplist[[i]]$cases)
    Toplist[[i]][, "cumdeaths"] <- cumsum(Toplist[[i]]$deaths)
    Toplist[[i]] <- Toplist[[i]][Toplist[[i]]$cumcases>0,]
    Toplist[[i]][, "datenorm"]  <- as.numeric(difftime(Toplist[[i]]$dateRep,
                                                       first(Toplist[[i]]$dateRep),
                                                       units="days"))    
}
```

\newpage
Plotting total number of cases vs time:
```{r cases-time, fig.height=5}
pchs <- rep(c(21,22,23,24,25), times=2)
maxcases <- max(totCases$x)
maxdays  <- as.numeric(difftime(yest, min(covid$dateRep)), units="days")

plot(Toplist[[1]]$datenorm, Toplist[[1]]$cumcases,
     main="Cumulative cases vs Time", xlab="Days from first case", ylab="Total cases", 
     col=cols[1], pch=pchs[1], cex=0.5,
     ylim=c(0,maxcases), xlim=c(0,maxdays),
     panel.first=grid(),
    )
for (i in c(2:10)) {
    points(Toplist[[i]]$datenorm, Toplist[[i]]$cumcases, 
           col=cols[i], cex=0.5, pch=pchs[i],
          )
}
par(xpd=TRUE)
legend("topleft", sort(TopgeoIds[[1]]), 
       col = cols,
       pch = pchs,
      )
```

\newpage
Plotting total number of deaths vs time:
```{r deaths-time, fig.height=5}
totDeaths <- aggregate(covid$deaths, by=list(covid$geoId), FUN=sum)
maxdeaths <- max(totDeaths$x)

plot(Toplist[[1]]$datenorm, Toplist[[1]]$cumdeaths, 
     main="Cumulative deaths vs Time", xlab="Days from first case", ylab="Total deaths", 
     col=cols[1], pch=pchs[1], cex=0.5, 
     ylim=c(0,maxdeaths), xlim=c(0,maxdays),
     panel.first=grid(),
    )
for (i in c(2:10)) {
    points(Toplist[[i]]$datenorm, Toplist[[i]]$cumdeaths, 
           col=cols[i], pch=pchs[i], cex=0.5,
          )
}
par(xpd=TRUE)
legend("topleft", sort(TopgeoIds[[1]]), 
       col = cols,
       pch = pchs,
      )
```

\newpage
Date of first recorded case in the top-ten countries:
```{r firstcase}
for (j in c(1:10)) {
    message(paste0(Toplist[[j]]$countryterritoryCode[1],": ",Toplist[[j]]$dateRep[1]))
} 
```
