---
title: "R Lab. - Exercise 2"
author: "Michele Guadagnini - Mt. 1230663"
date: "April 22, 2020"
header-includes:
   - \usepackage{multicol}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tibble)
library(dplyr)
```

## Exercise 0 - Practice with the discrete probability distributions in R

Practice with the discrete probability distributions in R
```{r practice, fig.height=6, fig.width=10, out.width='95%', fig.align='center'}
x <- 1:30
poisx <- dpois(x, 2)
binx  <- dbinom(x, 50, 0.2)
geomx <- dgeom(x, 0.2)
ppoisx <- ppois(x, 2)
pbinx  <- pbinom(x, 50, 0.2)
pgeomx <- pgeom(x, 0.2)
colors <- c("red", "blue", "darkgreen")
labels <- c("pois", "bin", "geom")

par(mfrow=c(1,2), oma=c(0,0,1,0))
plot(x, poisx, type="s", xlab="x value", ylab="Density", ylim=c(0,0.3),
     col=colors[1])
lines(x, binx, type="s", col=colors[2])
lines(x, geomx, type="s", col=colors[3])
legend("topright", inset=.05, labels, lwd=2, lty=c(1, 1, 1), col=colors)
plot(x, ppoisx, type="s", xlab="x value", ylab="Cumulative Density", ylim=c(0,1),
     col=colors[1])
lines(x, pbinx, type="s", col=colors[2])
lines(x, pgeomx, type="s", col=colors[3])
legend("bottomright", inset=.05, labels, lwd=2, lty=c(1, 1, 1), col=colors)
mtext("Comparison of Distributions", outer=TRUE, cex=1.5, line=-2)
```
\newpage

## Exercise 1 - Concentration of a contaminant in tap water
A set of measurements have been performed on the concentration of a contaminant in tap water with two methods. Evaluate the expected values, E[X], and the variance, Var(X), for both methods.
```{r ex1}
x  <- c(15.58, 15.9, 16 , 16.1, 16.2)
p1 <- c(0.15, 0.21, 0.35, 0.15, 0.14)
p2 <- c(0.14, 0.05, 0.64, 0.08, 0.09)

Expectation1 <- sum(x*p1)
Var1 <- sum(x*x*p1) - sum(x*p1)**2
Expectation2 <- sum(x*p2)
Var2 <- sum(x*x*p2) - sum(x*p2)**2
```
\begin{multicols}{2}
For method 1:\newline
the expected value is: `r Expectation1`\newline
the variance is: `r Var1`\newline
For method 2:\newline
the expected value is: `r Expectation2`\newline
the variance is: `r Var2`
\end{multicols}

## Exercise 2 - Waiting time at the doctor’s
The waiting time, in minutes, at the doctor’s is about 30 minutes, and the distribution follows an exponential pdf with rate 1/30

### A) simulate the waiting time for 50 people at the doctor’s office and plot the relative histogram
```{r doctor, fig.height=5, fig.width=9, out.width='95%'}
Npp <- 50
lambda <- 1/30
x <- 1:Npp
wt <- rexp(x, lambda)

hist(wt, col="red", density=40, xlim=c(0,max(wt)+30), ylim=c(0,35),
     ylab="Persons", xlab="Waiting time [min]", main="Simulated waiting times")
```

### B) what is the probability that a person will wait for less than 10 minutes?
```{r 10min}
theoProb10min <- pexp(10,lambda)
empProb10min  <- length(wt[wt<10])/length(wt)
```
Estimation from theoretical distribution: `r theoProb10min`\newline
Estimation from simulated data: `r empProb10min`

### C) evaluate the average waiting time from the simulated data and compare it with the expected value
```{r avgwt}
integrand <- function(z){z*dexp(z, lambda)}
integralWtavg <- integrate(integrand, 0, Inf)
theoricWtavg <- 1/lambda
empiricalWtavg <- mean(wt)
```
The estimation by integrating the distribution is: `r integralWtavg$value` min.\newline
The theoretical expected waiting time is: `r theoricWtavg` min.\newline
The waiting time estimated from simulation is: `r empiricalWtavg` min.

### D) what is the probability for waiting more than one hour before being received?
```{r 60min}
theoProb60min <- 1-pexp(60,lambda)
empProb60min  <- length(wt[wt>60])/length(wt)
```
The theoretical probability of waiting more then 60 minutes is: `r theoProb60min`\newline
The one estimated from simulation is: `r empProb60min`

\newpage
## Exercise 3 - Typos in a book
Suppose that on a book, on average, there is one typo error every three pages. If the number of errors follows a Poisson distribution, plot the pdf and cdf. 
```{r typos1, fig.height=5, fig.width=9, out.width='95%'}
Nmax <- 10
rate <- 1/3
x <- seq(0, Nmax, length=Nmax+1)
poispdf <- dpois(x, rate)
poiscdf <- ppois(x, rate)

par(mfrow=c(1,2))
plot(x, poispdf, type="b", xlab="# errors in a page", ylab="Probability",
     main="PDF", col=4, ylim=c(0,1))
plot(x, poiscdf, type='b', xlab="# errors in a page", ylab="Cumulative Probability",
     main="CDF", col=2, ylim=c(0,1))
```

Calculate the probability that there is at least one error on a specific page of the book.
```{r typos2}
Prob.1err <- 1-ppois(0,rate)   
```
The probability of finding one or more errors in a given page is `r Prob.1err`.

## Exercise 4 - Ace from a deck
We randomly draw cards from a deck of 52 cards, with replacement, until one ace is drawn. Calculate the probability that at least 10 draws are needed.

We need to exclude all the cases where we have less than 9 failures:
```{r ace}
p <- 4/52
minfails <- 8
Prob  <- 1 - pgeom(minfails, p)
```
The probability is: `r Prob`

\newpage
## Exercise 5 - Italian mayors database
Importing the dataframe containing all the italian mayors currently in charge:
```{r importdata}
dfmayors <- read.csv2("sindaciincarica.csv", skip=2)
names(dfmayors)
```

### A) Plot the gender distribution among the mayors.
```{r genddist, fig.height=7, fig.width=10, out.width='95%'}
genddist <- table(dfmayors$sesso)

bp<-barplot(genddist, beside=TRUE, col=c("mistyrose", "lightblue"),
            xlab="Gender", ylab="Counts", legend=c("F","M"), ylim=c(0,max(genddist)+1000),
            args.legend=list(x="topleft", inset=c(0.05,0)))
text(bp, genddist+200, labels=genddist, xpd=TRUE)
title( main="Italian mayors gender distribution" )
```

### B) Plot the number of towns grouped per province (codice provincia) and per region (codice regione).
```{r groupedtowns, fig.height=7, fig.width=10, out.width='95%'}
dfmayors <- tibble(dfmayors)

gdf <- group_split(dfmayors %>% group_by(dfmayors$codice_regione))
towntable <- vector("list", length(gdf))
for (i in 1:length(gdf)) {
    towntable[[i]] <- as.numeric(table(gdf[[i]]$codice_provincia,gdf[[i]]$codice_regione))
}
towntable <- as.data.frame(data.table::transpose(towntable), row.names=1:length(gdf))
color.ramp <- colorRampPalette(c("#0535F0", "#FFFFFF", "#FF0606"))(n=12)

barplot(t(as.matrix(towntable)), ylab="Region code", xlab="# Towns", 
        main="Italian towns grouped by region and province",
        border=F, las=1, horiz=T, xlim=c(0, 1600), col=color.ramp)
```

\newpage
### C) Plot a distributions of the age (years only) of the mayors.
```{r ages, fig.height=6, fig.width=10, out.width='95%'}
cyear <- as.numeric(format(Sys.Date(), "%Y"))
ages <- cyear - as.numeric(format(as.Date(dfmayors$data_nascita, format="%d/%m/%Y"),"%Y"))

hist(ages, col="green", xlim=c(0,100), density=40)
```

### D) Plot a distribution of the time the mayor is in charge. Since elections happen every 5 years, how many of them are going to complete their mandate this year? And how many in 2021?
```{r incharge, fig.height=6, fig.width=10, out.width='95%'}
cyear <- as.numeric(format(Sys.Date(), "%Y"))
elecyear <- format(as.Date(dfmayors$data_entrata_in_carica, format="%d/%m/%Y"),"%Y")
timeincharge <- cyear - as.numeric(elecyear)

breaks <- seq(0.5, 6.5, length=7)
htimeincharge <- hist(timeincharge, breaks=breaks, col="red", density=40,
           main="Histogram of Time in charge", xlab="Time in charge",
           ylim=c(0,4000))

outthisyear <- sum(htimeincharge$counts[5])
outnextyear <- sum(htimeincharge$counts[4])
total       <- sum(htimeincharge$counts)
```
The number of mayors completing their mandate this year are `r outthisyear` over a total number of `r total`. \newline
The number of mayors completing their mandate next year are `r outnextyear` over `r total`.