---
title: 'R Lab. - Exercise 5'
author: "Michele Guadagnini - Mt. 1230663"
date: "May 19, 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Exercise 1 - Launch of a new journal
A publishing company has recently launched a new journal. In order to determine how effective it is in reaching its possible audience, a market survey company selects a random sample of people from a possible target audience and interviews them. Out of 150 interviewed people, 29 have read the last issue of the journal.
\begin{itemize}
  \item[A)] What kind of distribution would you assume for y, the number of people that have seen the last issue of the journal?
  \item[B)] Assuming a uniform prior, what is the posterior distribution for y?
  \item[C)] Plot both posterior and likelihood distributions functions.
\end{itemize}
```{r ex1, fig.height=6, fig.width=8, out.width='80%', fig.align='center'}
## A ##  Assuming a binomial distribution as likelihood
likelihood <- function(p) { dbinom(29, size=150, prob=p) }
Ilike <- integrate(likelihood, 0, 1)$value
## B ##
prior <- function(p) { dunif(p, 0, 1) }
posterior <- function(p) { likelihood(p)*prior(p) }
Ipost <- integrate(posterior, 0, 1)$value
## C ##
p <- seq(0,1,length=2000)
plot(p,likelihood(p)/Ilike,type='l',col='blue',ylab='Density',lwd=3, main='Distributions')
lines(p, posterior(p)/Ipost, type='l', col='red', lty=2, lwd=3)
legend("topright",legend=c("Likelihood","Posterior"),col=c("blue", "red"),lty=1:1)
```

## Exercise 2 - New concert hall
Three students want to construct their prior probability about the proportion of residents that support the building of a new concert hall in their small town.
Anna thinks that her prior is a beta distribution with mean 0.2 and a standard deviation of 0.08. Benny moved only recently to this new town and therefore he does not have the slightest idea about it. Therefore he decides to use a uniform prior. Chris believes that his prior should have a trapezoidal shape.

### A) Draw and compare the three prior distributions.
```{r puntoa, fig.height=6, fig.width=8, out.width='95%', fig.align='center'}
mean <- 0.2
var  <- 0.08**2
anna.alpha <- mean*( (mean*(1-mean))/var -1 )
anna.beta  <- (1-mean)*( (mean*(1-mean))/var -1 )
anna.prior <- function(p) { dbeta(p, anna.alpha, anna.beta) }
benny.prior <- function(p) { dunif(p, 0, 1) }
chris.prior <- function(p){ ifelse(p>=0 & p<0.1, 20*p, 
                             ifelse(p>=0.1 & p<0.3, 2, 
                                   ifelse(p>=0.3 & p<0.5, 5-10*p, 0))) }
p <- seq(0,1,length=2000)
plot(p, anna.prior(p)/integrate(anna.prior,0,1)$value, type='l', col='red', lwd=2,
     ylab='Density',main='Prior distributions')
lines(p, benny.prior(p)/integrate(benny.prior,0,1)$value, type='l', col='blue', lwd=2)
lines(p, chris.prior(p)/integrate(chris.prior,0,1)$value, type='l', col='darkgreen',lwd=2)
legend("topright",legend=c("Anna's prior","Benny's prior","Chris' prior"),
       col=c("red","blue","darkgreen"),lty=1:1)
```

The next day the three students decide to interview a sample of 100 citizens of the small town, asking for their opinion. Out of the interviewed sample, 26 support the building of the new concert hall.

### B) Evaluate and draw the three posterior distributions.
```{r puntob, fig.height=7, fig.width=8, out.width='95%', fig.align='center'}
likelihood <- function(p) { dbinom(26, 100, p) }

anna.post <- function(p) { likelihood(p)*anna.prior(p) }
benny.post <- function(p) { likelihood(p)*benny.prior(p) }
chris.post <- function(p) { likelihood(p)*chris.prior(p) }

Ipost.anna <- integrate(anna.post,0,1)$value
Ipost.benny <- integrate(benny.post,0,1)$value
Ipost.chris <- integrate(chris.post,0,1)$value

plot(p, anna.post(p)/Ipost.anna, type='l', col='red',ylab='Density',
     main='Posterior distributions', lwd=1.5)
lines(p, benny.post(p)/Ipost.benny, col='blue', lwd=1.5)
lines(p, chris.post(p)/Ipost.chris, col='darkgreen', lwd=1.5)
legend("topright",legend=c("Anna's post.","Benny's post.","Chris' post."),
       col=c("red","blue","darkgreen"),lty=1:1)
```

### C) Give an estimate of the most probable value and the 95% credibility interval.
```{r puntoc, fig.height=6, fig.width=8, out.width='95%', fig.align='center'}
mode.anna <- p[which.max(anna.post(p))]
mode.benny <- p[which.max(benny.post(p))]
mode.chris <- p[which.max(chris.post(p))]

panna.post  <- Vectorize( function(p) { integrate(anna.post,0,p)$value/Ipost.anna } )
pbenny.post <- Vectorize( function(p) { integrate(benny.post,0,p)$value/Ipost.benny } )
pchris.post <- Vectorize( function(p) { integrate(chris.post,0,p)$value/Ipost.chris } )

low.anna <- p[max(which(panna.post(p)<=0.025))]
upp.anna <- p[min(which(panna.post(p)>=0.975))]

low.benny <- p[max(which(pbenny.post(p)<=0.025))]
upp.benny <- p[min(which(pbenny.post(p)>=0.975))]

low.chris <- p[max(which(pchris.post(p)<=0.025))]
upp.chris <- p[min(which(pchris.post(p)>=0.975))]
```

The most probable value and the $95\%$ credibility intervals are:
\begin{itemize}
  \item for Anna : `r round(mode.anna, 3)`, [`r round(low.anna, 3)`, `r round(upp.anna, 3)`]
  \item for Benny: `r round(mode.benny, 3)`, [`r round(low.benny, 3)`, `r round(upp.benny, 3)`]
  \item for Chris: `r round(mode.chris, 3)`, [`r round(low.chris, 3)`, `r round(upp.chris, 3)`]
\end{itemize}

\newpage

## Exercise 3 - Coin flipping
A coin is flipped n = 30 times with the following outcomes: \newline
T, T, T, T, T, H, T, T, H, H, T, T, H, H, H, T, H, T, H, T, H, H, T, H, T, H, T, H, H, H

### A) Assuming a flat prior, and a beta prior, plot the likelihood, prior and posterior distributions for the data set.
```{r puntoa3, fig.height=12, fig.width=8, out.width='95%', fig.align='center'}
flips <- c('T', 'T', 'T', 'T', 'T', 'H', 'T', 'T', 'H', 'H', 'T', 'T', 'H', 'H', 'H',
           'T',  'H', 'T', 'H', 'T', 'H', 'H', 'T', 'H', 'T', 'H', 'T', 'H', 'H', 'H')
flat.prior <- function(p) { dunif(p, 0, 1) }

mean <- 0.5
var  <- 0.01
alpha <- mean*( (mean*(1-mean))/var -1 )
beta  <- (1-mean)*( (mean*(1-mean))/var -1 )
beta.prior <- function(p) { dbeta(p, alpha, beta) }

r <- length(flips[flips=='H'])
n <- length(flips)
likelihood <- function(p) { dbinom(r, n, p) }

flat.post <- function(p) { likelihood(p)*flat.prior(p) }
beta.post <- function(p) { likelihood(p)*beta.prior(p) }

Ilike <- integrate(likelihood,0,1)$value
Iflat <- integrate(flat.post,0,1)$value
Ibeta <- integrate(beta.post,0,1)$value

par(mfrow=c(2,1))
p <- seq(0,1,length=2000)
plot(p, flat.post(p)/Iflat, col='red', type='l',ylab='Density',
     main='Distributions assuming flat prior',lwd=1.5)
lines(p, likelihood(p)/Ilike, col='blue',lty=2,lwd=1.5)
lines(p, flat.prior(p), col='darkgreen',lwd=1.5)
legend("topright",legend=c("Posterior","Likelihood","Prior"),
       col=c("red","blue","darkgreen"),lty=1:1)

plot(p, beta.post(p)/Ibeta, col='red', type='l',ylab='Density',
     main='Distributions assuming beta prior',lwd=1.5)
lines(p, likelihood(p)/Ilike, col='blue',lwd=1.5)
lines(p, beta.prior(p), col='darkgreen',lwd=1.5)
legend("topright",legend=c("Posterior","Likelihood","Prior"),
       col=c("red","blue","darkgreen"),lty=1:1)
```

### B) Evaluate the most probable value for the coin probability p and, integrating the posterior probability distribution, give an estimate for a 95% credibility interval.
```{r puntob3, fig.height=6, fig.width=8, out.width='95%', fig.align='center'}
mode.flat <- p[which.max(flat.post(p))]
mode.beta <- p[which.max(beta.post(p))]

pflat.post <- Vectorize( function(p) { integrate(flat.post,0,p)$value/Iflat } )
low.flat <- p[max(which(pflat.post(p)<=0.025))]
upp.flat <- p[min(which(pflat.post(p)>=0.975))]

pbeta.post <- Vectorize( function(p) { integrate(beta.post,0,p)$value/Ibeta } )
low.beta <- p[max(which(pbeta.post(p)<=0.025))]
upp.beta <- p[min(which(pbeta.post(p)>=0.975))]
```

The most probable value for the coin probability and the $95\%$ credibility intervals are:
\begin{itemize}
  \item with flat prior: `r round(mode.flat, 3)`, [`r round(low.flat, 3)`, `r round(upp.flat, 3)`]
  \item with beta prior: `r round(mode.beta, 3)`, [`r round(low.beta, 3)`, `r round(upp.beta, 3)`]
\end{itemize}

### C) Repeat the same analysis assuming a sequential analysis of the data. Show how the most probable value and the credibility interval change as a function of the number of coin tosses (i.e. from 1 to 30).
```{r puntoc3, fig.height=7, fig.width=8, out.width='95%', fig.align='center'}
mode.flat <- c()
low.flat <- c()
upp.flat <- c()
mode.beta <- c()
low.beta <- c()
upp.beta <- c()
for (i in 1:30) {
    current_flips <- flips[1:i]
    r <- length(current_flips[current_flips=='H'])
    n <- length(current_flips)

    #a beta distribution with alpha = beta = 1 is equivalent to a uniform distribution.
    #the sequential analysis is obtained updating the parameters (shape1, shape2)
    flat.post <- function(p) { dbeta(p, 1+r, 1+n-r) }
    
    #a beta dist. is a conjugate prior for a binomial likelihood, so in this case the
    # posterior distribution is still a beta distribution but with updated parameters.
    beta.post <- function(p) { dbeta(p, alpha+r, beta+n-r) }
    Iflat <- integrate(flat.post,0,1)$value
    Ibeta <- integrate(beta.post,0,1)$value
    
    mode.flat[i] <- p[which.max(flat.post(p))]
    pflat.post <- Vectorize ( function(p) { integrate(flat.post,0,p)$value/Iflat } )
    low.flat[i] <- p[max(which(pflat.post(p)<=0.025))]
    upp.flat[i] <- p[min(which(pflat.post(p)>=0.975))]

    mode.beta[i] <- p[which.max(beta.post(p))]
    pbeta.post <- Vectorize ( function(p) { integrate(beta.post,0,p)$value/Ibeta } )
    low.beta[i] <- p[max(which(pbeta.post(p)<=0.025))]
    upp.beta[i] <- p[min(which(pbeta.post(p)>=0.975))]
}

plot(1:30, mode.flat, pch=15, col='red', ylim=c(0,1),xlab='# coin flips',
     ylab='p',main='Flat prior',type='l')
points(1:30, low.flat, pch=15, col='blue',type='l')
points(1:30, upp.flat, pch=15, col='darkgreen',type='l')
legend("topright",legend=c("Mode","Lower limit","Upper limit"),
       col=c("red","blue","darkgreen"),pch=15)
grid()

plot(1:30, mode.beta, pch=15, col='red', ylim=c(0,1),xlab='# coin flips',
     ylab='p',main='Beta prior',type='l')
points(1:30, low.beta, pch=15, col='blue',type='l')
points(1:30, upp.beta, pch=15, col='darkgreen',type='l')
legend("topright",legend=c("Mode","Lower limit","Upper limit"),
       col=c("red","blue","darkgreen"),pch=15)
grid()
```

### D) Do you get a different result, by analyzing the data sequentially with respect to a one-step analysis (i.e. considering all the data as a whole)?

The most probable value for the coin probability and the $95\%$ credibility intervals at the end of the sequential analysis are:
\begin{itemize}
  \item with flat prior: `r round(mode.flat[30], 3)`, [`r round(low.flat[30], 3)`, `r round(upp.flat[30], 3)`]
  \item with beta prior: `r round(mode.beta[30], 3)`, [`r round(low.beta[30], 3)`, `r round(upp.beta[30], 3)`]
\end{itemize}

These values are equal to the ones obtained before in the one-step analysis.


