---
title: 'R Lab. - Exercise 6'
author: "Michele Guadagnini - Mt. 1230663"
date: "May 27, 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Exercise 1 - Radioactive source

The number of particles emitted by a radioactive source during a fixed interval of time ($\Delta$t =
10 s) follows a Poisson distribution on the parameter $\mu$. The number of particles observed during
consecutive time intervals is: 4, 1, 3, 1 and 3.

### A) suppose a uniform prior distribution for the parameter $\mu$:
\begin{itemize}
  \item determine and draw the posterior distribution for $\mu$, given the data
  \item evaluate mean, median and variance, both analytically and numerically in R
\end{itemize}

```{r puntoa1, fig.height=6, fig.width=8, out.width='95%', fig.align='center'}
dt <- 10 #seconds
outcomes <- c(4, 1, 3, 1, 3)

## A ##
mu <- seq(0, 10, len=1000)
prior <- 1
likelihood <- function(lam) { # likelihood as product of 5 independent measures
    k <- dpois(outcomes[1], lam)
    for (i in 2:length(outcomes)) {
        k <- k*dpois(outcomes[i], lam)
    }
    return (k)    
}
post.unif <- prior*likelihood(mu)
post.unif <- post.unif/sum(post.unif)

# numeric results
mean.num <- sum(mu*post.unif)
median.num <- mu[min( which(cumsum(post.unif) >= 0.5) )]
var.num  <- sum(mu*mu*post.unif) - mean.num**2
# analytic results
# posterior can be calculated also as: dgamma(mu, shape=13, rate=5)
shape <- sum(outcomes)+1
rate <- length(outcomes)
mean.th <- shape/rate
var.th  <- shape/(rate*rate)

# plotting the posterior distribution
plot(mu, post.unif/0.01, type='l', col='red', ylab="density",
     main="Posterior distribution with Uniform Prior")
```

The numerical results obtained in R are:
\begin{itemize}
  \item mean: `r round(mean.num, 3)`
  \item variance: `r round(var.num, 3)`
  \item median: `r round(median.num, 3)`
\end{itemize}

The analytical results obtained considering the posterior as a gamma distribution with proper parameters shape and rate are:
\begin{itemize}
  \item mean: `r round(mean.th, 3)`
  \item variance: `r round(var.th, 3)`
\end{itemize}

### B) suppose a Jeffrey’s prior for the parameter $\mu$:
\begin{itemize}
  \item determine and draw the posterior distribution for $\mu$, given the data
  \item evaluate mean, median and variance, both analytically and numerically in R
\end{itemize}

```{r puntob1, fig.height=6, fig.width=8, out.width='95%', fig.align='center'}
## B ##
mu <- seq(0.001, 10, len=1000)
prior <- function(x) { 1/sqrt(x) }
likelihood <- function(lam) { # likelihood as product of 5 independent measures
    k <- dpois(outcomes[1], lam)
    for (i in 2:length(outcomes)) {
        k <- k*dpois(outcomes[i], lam)
    }
    return (k)    
}
post.jeff <- prior(mu)*likelihood(mu)
post.jeff <- post.jeff/sum(post.jeff)

# numeric results
mean.num <- sum(mu*post.jeff)
median.num <- mu[min( which(cumsum(post.jeff) >= 0.5) )]
var.num  <- sum(mu*mu*post.jeff) - mean.num**2
# analytic results
# posterior can be calculated also as: dgamma(mu, shape=12.5, rate=5)
shape <- sum(outcomes)+1/2
rate <- length(outcomes)
mean.th <- shape/rate
var.th  <- shape/(rate*rate)

#plotting the posterior distribution
plot(mu, post.jeff/0.01, type='l', col='red', ylab="density", 
     main="Posterior distribution with Jeffrey's Prior")
```

\newpage

The numerical results obtained in R are:
\begin{itemize}
  \item mean: `r round(mean.num, 3)`
  \item variance: `r round(var.num, 3)`
  \item median: `r round(median.num, 3)`
\end{itemize}

The analytical results obtained considering the posterior as a gamma distribution with proper parameters shape and rate are:
\begin{itemize}
  \item mean: `r round(mean.th, 3)`
  \item variance: `r round(var.th, 3)`
\end{itemize}

### C) evaluate a 95% credibility interval for the results obtained with both priors. Compare the result with that obtained using a normal approximation for the posterior distribution, with the same mean and standard deviation
```{r puntoc1}
## C ##
low.unif <- mu[max(which(cumsum(post.unif)<=0.025))]
upp.unif <- mu[min(which(cumsum(post.unif)>=0.975))]

low.jeff <- mu[max(which(cumsum(post.jeff)<=0.025))]
upp.jeff <- mu[min(which(cumsum(post.jeff)>=0.975))]

normapprox <- pnorm(mu, mean.th, sqrt(var.th))
low.norm <- mu[max(which(normapprox<=0.025))]
upp.norm <- mu[min(which(normapprox>=0.975))]
```

The obtained 95% credibility intervals are:
\begin{itemize}
  \item with uniform prior: [`r round(low.unif, 3)`, `r round(upp.unif, 3)`]
  \item with Jeffrey's prior: [`r round(low.jeff, 3)`, `r round(upp.jeff, 3)`]
  \item for the normal approximation: [`r round(low.norm, 3)`, `r round(upp.norm, 3)`]
\end{itemize}

\newpage

## Exercise 2 - Lighthouse problem

Given the problem of the lighthouse discussed last week, study the case in which both the position along the shore ($\alpha$) and the distance out at sea ($\beta$) are unknown.

```{r puntoa2, fig.height=14, fig.width=10, out.width='95%', fig.align='center'}
# Generation of the data (flashes)
rx <- function(n, a, b) { 
    set.seed(13)
    x <- a + b*tan(runif(n,-pi/2,pi/2))
    return(x)
}

nsamples <- 100
alpha.true <- 1 #km
beta.true  <- 1 #km
xflashes <- rx(nsamples, alpha.true, beta.true)

# log of the posterior distribution
log.post <- function(flashes,a,b){
  logL <- 0.0
  for(x in flashes){
      logL <- logL + log(b/(pi*((x-a)**2 + b**2))) 
  }
  return(logL)
}

amin <- -4; amax <- 4; bmax <- 3
np <- 200
alphas <- seq(amin,amax, len=np)
betas  <- seq(   0,bmax, len=np)

ndata <- c(1,5,10,20,40,100)

par(mfrow=c(3,2))
for (i in ndata) {
    ff <- xflashes[1:i]
    func <- function(a,b) { log.post(ff, a, b) }

    log_values <- outer(alphas, betas, Vectorize(func))
    values <- exp(log_values)/( ((amax-amin)/np)*(bmax/np)*sum(exp(log_values)) )
    grid_values <- matrix(values, nrow=length(alphas), ncol=length(betas))
    
    contour(alphas, betas, grid_values, xlab="Alpha [km]", ylab="Beta [km]",
            main=paste("Lighthouse position probability with",i,"samples"), 
            col='darkorange', nlevels=5 )
    points(alpha.true, beta.true, pch=4, col="blue")
    max_ids <- c( which(grid_values==max(grid_values), arr.ind=TRUE) )
    points(alphas[max_ids[1]], betas[max_ids[2]], pch=4, col="red")
    legend("topleft", c("true position", "estimated pos."), col=c("blue","red"), pch=4)
    abline(v=alpha.true, h=beta.true, col="blue", lty=2)
    abline(v=alphas[max_ids[1]], h=betas[max_ids[2]], col="red", lty=2)
    grid()
}
```

\newpage

## Exercise 3 - Signal over Background

Given the Signal over Background discussed last week, analyze and discuss the following cases:

### A) vary the sampling resolution used to generate the data, keeping the same sampling range
```{r puntoa3, fig.height=5, fig.width=10, out.width='95%', fig.align='center'}
# signal function
signal <- function (x, a, b, x0, w, t) {
     t * (a*exp (-(x-x0)**2/(2*w**2)) + b)
}
# Log posterior
log.post <- function (d, x, a, b, x0, w, t) {
            if(a<0 || b <0) { return (-Inf )} # the effect of the prior
            sum( dpois(d, lambda = signal (x, a, b, x0, w, t), log=TRUE ))
}
# Parameters
x0 <- 0
dt <- 5 # exposure time in seconds
a.true <- 2
b.true <- 1
width <- 1

## A ##
ws <- c(0.1, 0.25, 1, 2, 3)

par(mfrow=c(1,2))
for (w in ws) {
    set.seed(241)
    xdat <- seq(from=-7*width, to=7*width, by=w*width)
    s.true <- signal(xdat, a.true, b.true, x0, width, dt)
    s.counts <- rpois(length(s.true), s.true)
    
    xps <- seq(min(xdat), max(xdat), len=1000)
    sps <- signal(xps, a.true, b.true, x0, width, dt)
    ymax <- max( c(max(sps), max(s.counts)) )
    
    plot(xps, sps, type='l', col='blue', ylim=c(0,ymax), xlab="x", ylab="counts",
         main=paste("Signal+Background with w =",w), xlim=c(-7,7), lwd=2)
    lines(xdat-w/2, s.counts, type='s', col='red' )
    
    a <- seq(0,4,len=100)
    b <- seq(0,2,len=100)
    delta_a <- 4/100
    delta_b <- 2/100
    # Computing log unnormalized posterior on a regular grid
    z <- matrix(data=NA , nrow= length(a), ncol= length(b))
    for (j in 1: length(a)) {
        for (k in 1: length(b)) {
            z[j,k] <- log.post(s.counts, xdat, a[j], b[k], x0, width, dt)
        }
    }
    # Normalizing posterior
    z.norm <- exp(z)/(delta_a*delta_b*sum(exp(z))) 
    
    # Plot normalized 2D posterior as contours
    contour(a, b, z.norm, nlevels = 4, labcex = 0.5,, xlab="amplitude, A", 
            ylab="background, B", col="darkorange", main=paste("2D posterior with w =",w))
    abline(v=a.true, h=b.true, col="darkgreen", lty=2)
    max_ids <- c( which(z.norm==max(z.norm), arr.ind=TRUE) )
    abline(v=a[max_ids[1]], h=b[max_ids[2]], col="red", lty=2)
    legend("topright", c("true position", "estimated pos."), 
           col=c("darkgreen","red"), lty=2)
}
```


### B) change the ratio A/B used to simulate the data (keeping both positive in accordance with the prior)
```{r puntob3, fig.height=12, fig.width=12, out.width='100%', fig.align='center'}
## B ##
# Parameters
x0 <- 0
dt <- 5     #exposure time in seconds
w  <- 0.5   #sampling resolution
width <- 1

as.true <- c(1,2,10,20)
bs.true <- c(1,1,1,1)

par(mfrow=c(2,2))
for (i in 1:4) {
    set.seed(2381)
    xdat <- seq(from=-7*width, to=7*width, by=w*width)
    s.true <- signal(xdat, as.true[i], bs.true[i], x0, width, dt)
    s.counts <- rpois(length(s.true), s.true)
    
    xps <- seq(min(xdat), max(xdat), len=1000)
    sps <- signal(xps, as.true[i], bs.true[i], x0, width, dt)
    ymax <- max( c(max(sps), max(s.counts)) )
    
    plot(xps, sps, type='l', col='blue', ylim=c(0,ymax), xlab="x", ylab="counts",
         main=paste("Signal+Background with A/B =",as.true[i]/bs.true[i]), 
         xlim=c(-7,7), lwd=2)
    lines(xdat-w/2, s.counts, type='s', col='red' )
    
    max.a <- as.true[i]+as.true[i]/2; max.b <- bs.true[i]+bs.true[i]/2
    a <- seq(0,max.a,len=100)
    b <- seq(0,max.b,len=100)
    delta_a <- max.a/100
    delta_b <- max.b/100
    # Computing log unnormalized posterior on a regular grid
    z <- matrix(data=NA , nrow= length(a), ncol= length(b))
    for (j in 1: length(a)) {
        for (k in 1: length(b)) {
            z[j,k] <- log.post(s.counts, xdat, a[j], b[k], x0, width, dt)
        }
    }
    # Normalizing posterior
    z.norm <- exp(z)/(delta_a*delta_b*sum(exp(z))) 
    
    # Plot normalized 2D posterior as contours
    contour(a, b, z.norm, nlevels = 4, labcex = 0.5,, xlab="amplitude, A",
             ylab="background, B", col="darkorange", 
             main=paste("2D posterior with A =",as.true[i],", B =",bs.true[i]))
    abline(v=as.true[i], h=bs.true[i], col="darkgreen", lty=2)
    max_ids <- c( which(z.norm==max(z.norm), arr.ind=TRUE) )
    abline(v=a[max_ids[1]], h=b[max_ids[2]], col="red", lty=2)
    legend("topleft", c("true position", "estimated pos."), 
           col=c("darkgreen","red"), lty=2)
}
```




